### 2010-07-23 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: uniform = vs. \mapsto: choose \mapsto
Desc@ICFP: uniform Desc and IDesc codes presentation
a bit of boiling in 3; abbreviated mu.de

While sigmas-of-sigmas can give a \emph{semantics} for the
sum-of-products structure in each node of the tree-like values in a
datatype, we need to account somehow for the recursive structure which
ties these nodes together. Not for the first time, we do this by
constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
are ubiquitous in dependently typed
programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
but here we seek to exploit them as the foundation of our notion
of datatypes.

@@

Descriptions are sequential structures, terminated by $\DUnit$,
indicating the empty tuple. To build sigmas-of-sigmas, we define a
$\SYMBDSigma$ code, interpreted as a $\Sigma$-type. To request a
recursive component, we have $\DIndx{\V{D}}$, where \(\V{D}\)
describes the rest of the node. These codes directly captures the
notion of sigmas-of-sigmas. An equivalent, more algebraic presentation
could be given, as illustrated in Section~\ref{sec:indexing-desc}.

\begin{figure}
\[\stk{
\begin{array}{l}
\Desc^n : \Set^{n+1}\\
\DUnit : \Desc^{n}\\
\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} : \Desc^{n}\\
\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} : \Desc^{n} \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{X}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc^{n} \To \Set^{n} \To \Set^{n}} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}\]

@@

You may have noticed that we are a little coy about this presentation,
writing of `implementing a signature' without clarifying how. A viable
approach would simply be to extend the theory with constants for the
constructors and an operator for \(\descop{\V{D}}\). However, in
Section~\ref{sec:desc-levitate}, you will see what we actually do.  In
the meantime, let us first gain some intuition for its use by
developing some examples.

@@

With a small change to this definition, we obtain the pattern functor
for lists:
%
\[\stk{
\ListD : \Set^n \To \Desc^n \\
\ListD \: \V{X} \mapsto
 \DSigma{\EnumT{\sqr{ \ListNil\: \SYMBListCons }}}
         {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
}\]
%
The $\SYMBNatSuc$ constructor is turned into a proper
$\SYMBListCons$, taking an argument in $\V{X}$ followed by a
recursive argument. This code describes the following functor:
%
\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of node-labelled binary trees:
%
\[\stk{
\TreeD : \Set^n \To \Desc^n \\
\begin{array}{@{}l@{}l}
\TreeD \: \V{X} \mapsto 
    \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
           { & \sqr{ \DUnit \quad
                     (\DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
\end{array}
}\]
%
Again, we are one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we add another. The
interpretation of this code corresponds to the following functor:
%
\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} 
          \V{Z} \Prod \V{X}  \Prod \V{Z}     \]

@@

From the examples above, we observe that datatypes are defined by a
$\SYMBDSigma$ whose first argument enumerates the constructors. We
call codes fitting this pattern \emph{tagged} descriptions. Again,
this is a clear reminder of the sum-of-products style. Every
description can be forced into this style with a singleton constructor
if necessary. We characterise tagged descriptions thus:
\[\stk{
 \TagDesc^n : \Set^{n+1} \\
 \TagDesc^n \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{\LAM{\_} \Desc^n})
\smallskip\\
\SYMBtoDesc : \TagDesc^n \To \Desc^n \\
\SYMBtoDesc \mapsto
\spl{\LAM{\V{E}}\LAM{\V{D}}
\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc^n)\:\V{D})}}
}\]
It is not a great stretch to imagine that the traditional datatype
declaration syntax might desugar to the \emph{definition} of a datatype
via a tagged description.

@@

So far, we have built pattern functors with our \(\Desc\) universe.
Being polynomial functors, they all admit a least fixpoint, which we
now construct by \emph{tying the knot}: the element type abstracted
by the functor is now instantiated recursively:
%
\[
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n}}
     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set^n}} \qquad
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n} \quad 
      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{\Mu{\M{D}}}}}
     {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
\]

@@

We can now build datatypes and their elements, e.g.:
\[\stk{
\Nat \mapsto \Bhab{\Mu{(\toDesc{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}})}}
                  {\Set^n}
\\
\Con{\sqr{\NatZero}}:\Nat \qquad
\Con{\sqr{\NatSuc{(\Bhab{n}{\Nat})}}}:\Nat
}\]

@@

However, iteration is inadequate for \emph{dependent} computation.
We need \emph{induction} to write functions whose type depends on inductive
data. Following \citet{benke:universe-generic-prog}, we adopt the following:
%
\[\stk{
\F{ind} : \stk{ \PITEL{\V{D}}{\Desc^n}
                \PI{\V{P}}{\Mu{\V{D}} \To \Set^k}         \\
               (\PI{\V{d}}{\descop{\V{D}}{\Mu{\V{D}}}}       
                \All{\V{D}}{\Mu{\V{D}}}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
               \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
} \\
\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) = 
    \V{m}\, \V{d}\, (\all{\V{D}}{\Mu{\V{D}}}{\V{P}}
                           {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
                           {\V{d}})
}\]
%
Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set^k}$ holds for every subobject $\V{x}:\V{X}$
in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
`dependent map', applying some
\(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
contained in \(\V{d}\). The full definition (including an extra case,
introduced shortly) is presented in
Figure~\ref{fig:all-predicates}.\footnote{To pass the termination
  checker, we had to inline the definition of $\SYMBall$ into
  $\SYMBind$ in our Agda model. A simulation argument shows that the
  definition presented here terminates if the inlined version
  does. Hence, although not directly structural, this definition is
  indeed terminating.} Note that $\SYMBind$ is our first generic
operation over descriptions, albeit a hardwired operator.  Any
datatype we define automatically comes with an induction principle.

We note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
fixpoints, and we have indeed implemented coinductive types this way, but
that is a story for another time.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAll : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \Set^k 
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
    \Unit \\
\All{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{s}}{\V{d}}{}} &
    \All{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{d}} \\
\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{x}}{\V{d}}{}} &
    \TIMES{\V{P}\: \V{x}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{d}}{}} &
    \TIMES{(\PI{\V{h}}{\V{H}} \V{P}\: (\V{f}\: \V{h}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBall : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PITEL{\V{p}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\all{\DUnit}{& \V{X}}{\V{P}}{\V{p}}{\Void} &
    \Void \\
\all{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \all{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{x}}{\V{d}}{}} &
    \pair{\V{p}\: \V{x}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{} \\
\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{f}}{\V{d}}{}} &
    \pair{\LAM{\V{h}} \V{p}\: (\V{f}\: \V{h})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{}
\end{array}
\end{array}
}
\]

@@

We have now enough machinery to build and manipulate inductive
types at a low level. Let us now apply cosmetic surgery
to the syntactic overhead. We extend type checking of
expressions:
%
\[
\Rule{\Gamma\Vdash\propag{\push{\etag{c}}{\EnumT{\M{E}}}}{n} \quad
\Gamma\Vdash  
\propag{
  \push{\sqr{\vec{t}}}
   {\descop{\M{D}\: n}{\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}}
        {\M{t'}}}
     {\Gamma\Vdash\propag{\push{\etag{c}\: \vec{\M{t}}}
                   {\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}
             {\Con{\pair{n}{\M{t'}}{}}}}
\]
%
Here $\etag{c}\:\vec{t}$ denotes a tag `applied' to a sequence
of arguments, and \(\sqr{\vec{t}}\) that sequence's repackaging as
a right-nested tuple. Now we can just write data directly.
\[
\NatZero:\Nat\qquad \NatSuc{(\Bhab{n}{\Nat})}:\Nat
\]
Once again, the type explains the legible presentation, as well as
the low-level representation.

We may also simplify appeals to induction by type propagation, as we
have done with functions from pairs and enumerations.
\[
\Rule{\Gamma\Vdash\stk{
\propag{\push{\M{f}}
             {\PI{\V{d}}{\descop{\M{D}}{\Mu{\M{D}}}}       
                \All{\M{D}}{\Mu{\M{D}}}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
       {\M{f'}}}}
{\Gamma\Vdash
\propag{\push{\sind{\M{f}}}
             {\PIS{\Bhab{\x}{\Mu{\M{D}}}}\M{P} }}
       {\F{ind}\:\M{D}\:(\PLAM{\x}{\Mu{\M{D}}}\M{P})\:\M{f'}}}
\]
This abbreviation is no substitute for the dependent pattern
matching to which we are entitled in a high-level language built on top
of this theory~\cite{goguen:pattern-matching}, but it does at least make
`assembly language' programming mercifully brief, if hieroglyphic.
\[\stk{
\F{plus}:\Nat\To\Nat\To\Nat \\
\F{plus}\mapsto\sind{\spl{\sqr{(\LAM{\_}\LAM{\_}\LAM{\V{y}}\V{y})
  \quad(\LAM{\_}\spl{\LAM{\V{h}}\LAM{\_}\LAM{\V{y}}
    \NatSuc{(\V{h}\:\V{y})}})}}}
}\]

>>>>

While sigmas-of-sigmas can give a \emph{semantics} for the
sum-of-products structure in each node of the tree-like values in a
datatype, we need to account for the recursive structure which
ties these nodes together. As ever, we do this by
constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
are ubiquitous in dependently typed
programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
but here we take them as the foundation of our notion
of datatypes.

@@

Descriptions are sequential structures ending in $\DUnit$,
indicating the empty tuple. To build sigmas-of-sigmas, we provide a
$\SYMBDSigma$ code, interpreted as a $\Sigma$-type. To request a
recursive component, we have $\DIndx{\V{D}}$, where \(\V{D}\)
describes the rest of the node. These codes give us
sigmas-of-sigmas with recursive places. An equivalent, more algebraic presentation
could be given, as illustrated in Section~\ref{sec:indexing-desc}.

\begin{figure}
\[\stk{
\begin{array}{@{}ll}
\Desc^n 
    &: \Set^{n+1}\\
\DUnit 
    &: \Desc^{n}\\
\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} 
    &: \Desc^{n}\\
\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} 
    &: \Desc^{n} \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{X}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc^{n} \To \Set^{n} \To \Set^{n}} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}\]

@@

We admit to being a little coy,
writing of `implementing a signature' without clarifying how. A viable
approach would simply be to extend the theory with constants for the
constructors and an operator for \(\descop{\V{D}}\). In
Section~\ref{sec:desc-levitate}, you will see what we do instead.
Meanwhile, let us gain some intuition by developing examples.

@@

The pattern functor
for lists needs but a small change:
%
\[\stk{
\ListD : \Set^n \To \Desc^n \\
\ListD \: \V{X} \mapsto
 \DSigma{\EnumT{\sqr{ \ListNil\: \SYMBListCons }}}
         {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
}\]
%
The $\SYMBNatSuc$ constructor becomes
$\SYMBListCons$, taking an $\V{X}$ followed by a
recursive argument. This code describes the following functor:
%
\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

@@

Of course, we are not limited to one recursive argument. Here are
the node-labelled binary trees:
%
\[\stk{
\TreeD : \Set^n \To \Desc^n \\
\begin{array}{@{}l@{}l}
\TreeD \: \V{X} \mapsto 
    \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
           { & \sqr{ \DUnit \quad
                     (\DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
\end{array}
}\]
%
Again, we are one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we add another. The
interpretation of this code corresponds to the following functor:
%
\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} 
          \V{Z} \Prod \V{X}  \Prod \V{Z}     \]

@@

From the examples above, we observe that datatypes are defined by a
$\SYMBDSigma$ whose first argument enumerates the constructors. We
call codes fitting this pattern \emph{tagged} descriptions. Again,
this is a clear reminder of the sum-of-products style. Any
description can be forced into this style with a singleton constructor
set. We characterise tagged descriptions thus:
\[\stk{
 \TagDesc^n : \Set^{n+1} \\
 \TagDesc^n \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{\LAM{\_} \Desc^n})
\smallskip\\
\SYMBtoDesc : \TagDesc^n \To \Desc^n \\
\SYMBtoDesc \mapsto
\spl{\LAM{\V{E}}\LAM{\V{D}}
\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc^n)\:\V{D})}}
}\]
It is not such a stretch to expect that the familiar datatype
declaration might desugar to the \emph{definitions} of a tagged description.

@@

\newcommand{\mude}[1]{\D{\(\mu^{\!+}\)}\:#1}

So far, we have built pattern functors with our \(\Desc\) universe.
Being polynomial functors, they all admit a least fixpoint, which we
now construct by \emph{tying the knot}: the element type abstracted
by the functor is now instantiated recursively:
%
\[
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n}}
     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set^n}} \qquad
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n} \quad 
      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}}
     {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
\]
Tagged descriptions are very common, so we abbreviate:
\[
  \mude{}:\TagDesc^n \To \Set^n \qquad \mude{\V{T}}\mapsto\Mu{(\toDesc{\V{T}})}
\]

@@

We can now build datatypes and their elements, e.g.:
\[\stk{
\Nat \mapsto \Bhab{\mude{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}}}
                  {\Set^n}
\\
\Con{\sqr{\NatZero}}:\Nat \qquad
\Con{\sqr{\NatSuc{(\Bhab{n}{\Nat})}}}:\Nat
}\]

@@

However, iteration is inadequate for \emph{dependent} computation.
We need \emph{induction} to write functions whose type depends on inductive
data. Following \citet{benke:universe-generic-prog}, we adopt the following:
%
\[\stk{
\F{ind} : \stk{ \PITEL{\V{D}}{\Desc^n}
                \PI{\V{P}}{\Mu{\V{D}} \To \Set^k}         \\
               (\PI{\V{d}}{\descop{\V{D}}{(\Mu{\V{D}})}}       
                \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
               \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
} \\
\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) \mapsto 
    \V{m}\, \V{d}\, (\all{\V{D}}{(\Mu{\V{D}})}{\V{P}}
                           {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
                           {\V{d}})
}\]
%
Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set^k}$ holds for every subobject $\V{x}:\V{X}$
in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
`dependent map', applying some
\(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
contained in \(\V{d}\). The definition (including an extra case,
introduced soon) is in
Figure~\ref{fig:all-predicates}.\footnote{To pass the termination
  checker, we had to inline the definition of $\SYMBall$ into
  $\SYMBind$ in our Agda model. A simulation argument shows that the
  definition presented here terminates if the inlined version
  does. Hence, although not directly structural, this definition is
  indeed terminating.} So, $\SYMBind$ is our first
operation generic over descriptions, albeit hardwired.  Any
datatype we define comes with induction.

Note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
fixpoints: we have indeed implemented coinductive types this way, but
that is another story.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAll : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \Set^k 
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
    \Unit \\
\All{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{s}}{\V{d}}{}} &
    \All{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{d}} \\
\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{x}}{\V{d}}{}} &
    \TIMES{\V{P}\: \V{x}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{d}}{}} &
    \TIMES{(\PI{\V{h}}{\V{H}} \V{P}\: (\V{f}\: \V{h}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBall : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PITEL{\V{p}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\all{\DUnit}{& \V{X}}{\V{P}}{\V{p}}{\Void} &
    \Void \\
\all{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \all{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{x}}{\V{d}}{}} &
    \pair{\V{p}\: \V{x}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{} \\
\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{f}}{\V{d}}{}} &
    \pair{\LAM{\V{h}} \V{p}\: (\V{f}\: \V{h})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{}
\end{array}
\end{array}
}
\]

@@

We now have low level machinery to build and manipulate inductive
types. Let us apply cosmetic surgery
to reduce the syntactic overhead. We extend type checking of
expressions:
%
\[
\Rule{\Gamma\Vdash\propag{\push{\etag{c}}{\EnumT{\M{E}}}}{n} \quad
\Gamma\Vdash  
\propag{
  \push{\sqr{\vec{t}}}
   {\descop{\M{D}\: n}{(\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})})}}}
        {\M{t'}}}
     {\Gamma\Vdash\propag{\push{\etag{c}\: \vec{\M{t}}}
                   {\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}
             {\Con{\pair{n}{\M{t'}}{}}}}
\]
%
Here $\etag{c}\:\vec{t}$ denotes a tag `applied' to a sequence
of arguments, and \(\sqr{\vec{t}}\) that sequence's repackaging as
a right-nested tuple. Now we can just write data directly.
\[
\NatZero:\Nat\qquad \NatSuc{(\Bhab{n}{\Nat})}:\Nat
\]
Once again, the type explains the legible presentation, as well as
the low-level representation.

We may also simplify appeals to induction by type propagation, as we
have done with functions from pairs and enumerations.
\[
\Rule{\Gamma\Vdash\stk{
\propag{\push{\M{f}}
             {\PI{\V{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}       
                \All{\M{D}}{(\Mu{\M{D}})}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
       {\M{f'}}}}
{\Gamma\Vdash
\propag{\push{\sind{\M{f}}}
             {\PIS{\Bhab{\x}{\Mu{\M{D}}}}\M{P} }}
       {\F{ind}\:\M{D}\:(\PLAM{\x}{\Mu{\M{D}}}\M{P})\:\M{f'}}}
\]
This abbreviation is no substitute for the dependent pattern
matching to which we are entitled in a high-level language built on top
of this theory~\cite{goguen:pattern-matching}, but it does at least make
`assembly language' programming mercifully brief, if hieroglyphic.
\[\stk{
\F{plus}:\Nat\To\Nat\To\Nat \\
\F{plus}\mapsto\sind{\spl{\sqr{(\LAM{\_}\LAM{\_}\LAM{\V{y}}\V{y})
  \quad(\LAM{\_}\spl{\LAM{\V{h}}\LAM{\_}\LAM{\V{y}}
    \NatSuc{(\V{h}\:\V{y})}})}}}
}\]
